Using VariantSpark to learn population structure from genomic profiles

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
made all the lines shorter (don't you guys ahve a carriage return?). make all the TODOs into marginpars

Aiding this task, we developed \variantSpark~\cite{OBrien}, a \kMeans{} approach using {\sc Spark}'s machine learning library, {\sc MLlib}, which is capable of clustering variants from whole genome sequencing data. 
We demonstrate its use by clustering samples from the 1000 Genomes Project~\cite{1KG2012}.
%TODO add value of the kmeans method on the full dataset (the equivalent of the ARI=0.84)
% Denis: correct/total happened to be 0.84 too ? or are you quoting here the ARI ? I want to have the same accuracy measure as reported by bdgenomics for all comparisons in this paper, also it goes between 0 and 100 right ?

@@

In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population association and compare our results to the ones obtained using deep learning. 
%TODO Rob can you think of a better description of what it does
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables. 
It is therefore able to adjust the weight of the individual genotypes to increase the accuracy of the multi-class prediction.

%TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}. 
We also investigate more complex supervised approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal the rest seem to jointly predict the complex metabolic phenotype. 
We therefore investigate whether ethnicity is a similarly complex phenotype that requires higher order dependencies between features to be taken into account. %, as done by RF and DL.
%TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect to the multi variable interactions and RF benefit of being hard to overfit).
%Using conventional machine learning analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model over such a high-dimensional feature space~\cite{Chen2012}. 
%As a result machine learning methods such as random forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make the approach robust against overfitting issues~\cite{Breiman2001}. 
However, training a supervised learning method on such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.

@@

%TODO check accuracy of titles
\section{Results}

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.
%TODO elaborate on the error
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40 Million variables, due to XXX. 
We therefore use only chromosome 22 for the comparison in this section.
We train a RF and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all super-populations.
We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
We also perform 5 runs of \kMeans{} with 4 clusters and report the mean accuracy of the resulting models along with the standard error.

%TODO elaborate on accuracy measures
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately (XX\%, ARI=Y), see table~\ref{RF}.
RF achieves a higher accuracy (XX\%, ARI=Y) but is outperformed by the simple logistic regression (XX\%, ARI=Y).
This indicates that a multivariate approach may not be needed to accurately fit ethnicity.

@@

%TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our paper)
\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than RF by again comparing it against \kMeans{} and logistic regression.  
We compare against the reported accuracy of the deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}. 
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW (Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and Scotland).

@@

%TODO[say that this may be a too easy dataset to benchmark the performance differences between kmeans and RF (hence the full dataset needed)]

@@

%TODO [Discuss supervised vs unsupervised]
%TODO can you drill down to the population clusters and see if and where the best performing method still makes mistakes - this will help in the conclusion to argue that actually the labels are wrong rather than the method not being able to capture the structure (e.g. say if AMR are mixed in with GBR TSI)

@@

%TODO [write up the observations and argue why one is less resource hungry than the other one]

@@

\bibliographystyle{plain}
%TODO you need to find a way of shortening the Locket paper author list
\bibliography{recomb}

>>>>

Aiding this task, we developed \variantSpark~\cite{OBrien}, a \kMeans{} approach using {\sc Spark}'s machine learning
library, {\sc MLlib}, which is capable of clustering variants from whole genome sequencing data.  We demonstrate its use
by clustering samples from the 1000 Genomes Project~\cite{1KG2012}.  \marginpar{TODO add value of the kmeans method on
  the full dataset (the equivalent of the ARI=0.84) Denis: correct/total happened to be 0.84 too ? or are you quoting
  here the ARI ? I want to have the same accuracy measure as reported by bdgenomics for all comparisons in this paper,
  also it goes between 0 and 100 right ?}

@@

In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
association and compare our results to the ones obtained using deep learning.
\marginpar{TODO Rob can you think of a better description of what it does}
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
by fitting a linear combination of the feature variables.  It is therefore able to adjust the weight of the individual
genotypes to increase the accuracy of the multi-class prediction.

\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  We also investigate more complex supervised
approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal
the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
similarly complex phenotype that requires higher order dependencies between features to be taken into
account. %, as done by RF and DL.
\marginpar{TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect
  to the multi variable interactions and RF benefit of being hard to overfit).  Using conventional machine learning
  analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model
  over such a high-dimensional feature space~\cite{Chen2012}.  As a result machine learning methods such as random
  forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make
  the approach robust against overfitting issues~\cite{Breiman2001}.  However, training a supervised learning method on
  such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel
  approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models
  based on information from the full dataset.}

@@

\marginpar{TODO check accuracy of titles}
\section{Results}

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
\marginpar{TODO elaborate on the error} 
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to XXX.  We therefore use only chromosome 22 for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds
along with the standard error.  We also perform 5 runs of \kMeans{} with 4 clusters and report the mean accuracy of the
resulting models along with the standard error.

\marginpar{TODO elaborate on accuracy measures}
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately (XX\%, ARI=Y), see table~\ref{RF}.
RF achieves a higher accuracy (XX\%, ARI=Y) but is outperformed by the simple logistic regression (XX\%, ARI=Y).
This indicates that a multivariate approach may not be needed to accurately fit ethnicity.

@@

\marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
  paper)}
\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and
Scotland).

@@

\marginpar{TODO[say that this may be a too easy dataset to benchmark the performance differences between kmeans and RF
  (hence the full dataset needed)]}

@@

\marginpar{TODO [Discuss supervised vs unsupervised]} \marginpar{TODO can you drill down to the population clusters and
  see if and where the best performing method still makes mistakes - this will help in the conclusion to argue that
  actually the labels are wrong rather than the method not being able to capture the structure (e.g. say if AMR are
  mixed in with GBR TSI)}

@@

\marginpar{TODO [write up the observations and argue why one is less resource hungry than the other one]}

@@

\bibliographystyle{plain}
\marginpar{TODO you need to find a way of shortening the Locket paper author list}
\bibliography{recomb}
